### 2012-10-30 paper_sean/conclusion.tex
Rewrote Results with Sean, plus other changes. All remaining needed changes are reflected in edit blocks.

% !TEX root =main.tex
HyperNEAT has proven to be an effective encoding in gait evolution with an ability to find gait regularities without human intervention~\cite{clune2009evolving,clune2011performance,yos:clune}. Glette et al. 2012~\cite{glette} showed the benefits of a physical simulator by evolving, with a simple genetic algorithm, a gait that outperformed those HyperNEAT evolved in hardware.
In this study, we took the two -- HyperNEAT and a physical simulator -- and showed that HyperNEAT's performance significantly increases with longer evolution periods. The gaits produced in this study outperformed the previous best gait on the QuadraTot by 54.3\% in simulation and by 5.4\% in real life trials. This study also showed that HyperNEAT will exploit the characteristics of the problem, such as frequency and regularity, in order to maximize its fitness.
%In simulation, HyperNEAT gaits outperformed Genetic Algorithm gaits~\cite{glette} by 54.3\% and by 5.4\% in the real life trials.

>>>>

% !TEX root =main.tex

\edit{Jeff: use me} Overall, these results provide important confirmation that the HyperNEAT generative encoding can evolve state-of-the-art results for challenging engineering problems, such as evolving gaits for real, legged robots. The results also reaffirm the benefits of using simulators when solving real-world challenges with evolutionary algorithms.

\edit{Last: check and edit this section.}

HyperNEAT has proven to be an effective encoding in gait evolution with an ability to find gait regularities without human intervention~\cite{clune2009evolving,clune2011performance,yos:clune}. Glette et al. 2012~\cite{glette} showed the benefits of a physical simulator by evolving, with a simple genetic algorithm, a gait that outperformed those HyperNEAT evolved in hardware.
In this study, we took the two -- HyperNEAT and a physical simulator -- and showed that HyperNEAT's performance significantly increases with longer evolution periods. The gaits produced in this study outperformed the previous best gait on the QuadraTot by 54.3\% in simulation and by 5.4\% in real life trials. This study also showed that HyperNEAT will exploit the characteristics of the problem, such as frequency and regularity, in order to maximize its fitness.
%In simulation, HyperNEAT gaits outperformed Genetic Algorithm gaits~\cite{glette} by 54.3\% and by 5.4\% in the real life trials.

### 2012-10-30 paper_sean/introduction.tex

Robots that can move around on legs can operate in a much wider range of environments than their wheeled counterparts. A downside is that designing gaits for legged robots is a difficult and time-consuming process for human engineers~\cite{strom1999legged,wettergreen1992gait}, and must be repeated every time a new robot is created or the design of an existing robot is changed~\cite{hornby2005autonomous}. 
Because of these difficulties in hand-designing gaits, scientists have investigated automatically producing gaits with machine learning or evolutionary algorithms, an approach that often produces better gaits than those created by human engineers~\cite{valsalam:mii,kohl:stone,hornby2005autonomous,hornby2003generative,yos:clune}. It has often been found that both machine-created and human-created gaits perform better if they are \emph{regular}, in that they have coordinated movements, such as left-right symmetry or front-back symmetry~\cite{valsalam:mii,clune2011performance,clune2009evolving,clune2009sensitivity}.
However, in most previous works, the experimenter has had to manually decide the regularities of the evolved gaits~\cite{valsalam:mii,tellez,beer,raibert}. 
Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance.

Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding~\cite{stanley2009hypercube} can automatically generate a variety of regular gaits and that it outperforms direct encoding controls on this task~\cite{clune2009evolving,clune2011performance}.
However, that work only verified these claims in simulation. 
A follow-up paper by Yosinski et al. evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform and found that HyperNEAT's gaits outperformed manually designed parameterized learning algorithms, but still did not produce impressive, natural gaits~\cite{yos:clune}. 
A further follow-up study built a simulator for the QuadraTot robot platform to test whether the inclusion of a simulator would improve results and found that it did:  when gaits were evolved in this simulator with a simple direct encoding and then transferred to the QuadraTot robot, the resulting gaits were faster than the gaits produced by evolving gaits with HyperNEAT directly on the robot~\cite{glette}. 
The simulator helped because it afforded much larger population sizes and more generations than were possible when evolving directly in hardware. Specifically, the simulator allowed 333 times as many evaluations per run as HyperNEAT had in hardware (60000 vs. 180)~\cite{yos:clune,glette}.

The work with the simulator (Glette et al. 2012~\cite{glette}) evolved gaits with a simple encoding that was manually constrained to produce specific regularities. The success of that work raises the question of whether the performance gains were due to the use of a simulator, or the use of a simple, manually-constrained encoding. We hypothesized that HyperNEAT, which has been previously shown to automatically discover complex regularities to produce high-performing gaits~\cite{clune2011performance,yos:clune}, would outperform the simpler encoding from Glette et al. 2012~\cite{glette} if HyperNEAT is combined with a simulator. 
Here we test that hypothesis by evolving gaits with HyperNEAT in the same simulator from Glette et al. and then transferring those gaits to the QuadraTot robot.  Our experiments confirm the hypothesis: gaits evolved with HyperNEAT and then transferred to reality were the highest performing observed to date on the QuadraTot robot platform.

>>>>

Legged robots can operate in a much wider range of environments than their wheeled counterparts. However, designing gaits for legged robots is a difficult and time-consuming process for human engineers~\cite{strom1999legged,wettergreen1992gait}, and must be repeated every time a robot is created or modified~\cite{hornby2005autonomous}. 
Scientists thus investigate how to automatically produce gaits via machine learning and evolutionary algorithms, which often produces better gaits than those created by human engineers~\cite{valsalam:mii,kohl:stone,hornby2005autonomous,hornby2003generative,yos:clune}. While it has been shown that gaits perform better if they are \emph{regular}, in that they have coordinated movements, such as left-right symmetry or front-back symmetry~\cite{valsalam:mii,clune2011performance,clune2009evolving,clune2009sensitivity}, experimenters usually have to manually decide and specify these regularities~\cite{valsalam:mii,tellez,beer,raibert}. 
Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance.

Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding~\cite{stanley2009hypercube} can automatically generate a variety of regular gaits that outperform gaits evolved with direct encodings~\cite{clune2009evolving,clune2011performance}.
However, that work only verified these claims in simulation. 
Yosinski et al. evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform~(\figref{quadratotWhiteBooties}). They found that HyperNEAT's gaits outperformed manually designed, parameterized learning algorithms, but still did not produce impressive, natural gaits~\cite{yos:clune}. 
A follow-up study built a simulator for QuadraTot to test whether the inclusion of a simulator would improve results and found that it did:  when gaits were evolved in this simulator with a simple direct encoding and then transferred to the QuadraTot robot, the resulting gaits were faster than the gaits produced by evolving gaits with HyperNEAT directly on the robot~\cite{glette}. 
The simulator helped because it afforded much larger population sizes and more generations than were possible when evolving directly in hardware, resulting in 333 times more evaluations per run (60000 vs. 180)~\cite{yos:clune,glette}.

The work with the simulator (Glette et al. 2012) evolved gaits with a simple encoding that was manually constrained to produce specific regularities. The success of that work raises the question of whether the performance gains were due to the added benefit of a simulator or the use of a simple, hand-designed encoding. We hypothesized that HyperNEAT, which has been previously shown to automatically discover complex regularities to produce high-performing gaits~\cite{clune2011performance,yos:clune}, would outperform the simpler encoding from Glette et al. if HyperNEAT is combined with a simulator. 
Here we test that hypothesis by evolving gaits with HyperNEAT in the same simulator from Glette et al. and then transferring those gaits to the QuadraTot robot.  Our experiments confirmed the hypothesis: with the simulator, HyperNEAT evolved the highest-performing gaits observed to date for the QuadraTot robot platform.

### 2012-10-30 paper_sean/methods.tex

\subsection{Methods}

We evolved gaits first in simulation and then transferred them to the physical QuadraTot robot, as in Glette et al. 2012~\cite{glette}. 
Each run had a population size of 200 and lasted 200 generations. 20 runs were performed, each of which was identical except for the seed provided to the random number generator, which affected stochastic events such as mutation. 
Each gait was evaluated for 12 seconds \edit{JBY: how? As in fig 4 of \cite{yos:clune}, or another way?}, both in simulation and in reality. 
All HyperNEAT parameters were identical to those in Yosinski et al. 2011~\cite{yos:clune} except for the frequency of the sine wave input to the ANN, which was lowered from 0.0265Hz to 0.004Hz to decrease the number of servo shutdowns due to high-frequency gaits \edit{JBY: what? why?}.

%\begin{equation}
%sin(2\pi t/250)\pi
%\end{equation}

%from
%
%\begin{equation}
%sin(t/6)\pi
%\end{equation}
%

 %Sine frequency of 0.004 was used because we experimentally found that HyperNEAT was able to produce natural and high-performing gaits with low frequencies with this frequency. 
We also punished high-frequency gaits during evolving in simulation, which was effective at reducing the number of servo shutdowns observed when evolved gaits were transferred to the robot. 
If the average frequency of each leg \edit{JBY: any leg or all legs?} for the gait was higher than a threshold frequency (1.67 Hz), the fitness was punished by exponentially reducing the distance travelled by the robot:

\begin{equation}
\mathrm{distance} = \mathrm{distance} \cdot e^{(\mathrm{freq}_{\mathrm{thr}}-\mathrm{freq})} \\
\end{equation}

\noindent where \emph{$\mathrm{distance}$} is the Euclidean distance the center of mass of the robot travelled during the gait duration. Following Clune et al. 2009~\cite{clune2009evolving}, the fitness equation was

\begin{equation}
\mathrm{fitness} = 2^{\mathrm{distance}^{2}}
\end{equation}

After evolving the gaits in simulation, the champion gait of the last generation of each of the 20 runs was transferred onto the real robot and evaluated.
The robot's position was measured at the beginning and the end of each run, starting from the stationary pose and ending with the stationary pose. 
If a gait was obviously affected by servo shutdowns, the gait was tried again after letting the servos cool down.

>>>>

\subsubsection{Evolutionary Process and Parameters:}

Each run had a population size of 200 and lasted 200 generations. We performed twenty HyperNEAT runs that differed only in the seed provided to the random number generator, which affected stochastic events such as mutation. To make a statistical comparison to the encoding from Glette et al., we conducted 19 runs using the original code to supplement the one run performed for that paper. 
Each gait was evaluated for fourteen seconds in reality, with interpolation from and to a stationary position in the first and last two seconds, respectively, as done in Yosinski et al. The result was effectively 12 seconds of full-speed
motion. In simulation gaits were evaluated for 12 seconds. 
All HyperNEAT parameters were identical to those in Yosinski et al. except for the frequency of the sine wave input to the ANN, which was lowered from 4.2Hz to 0.64Hz to reduce servo shutdowns. 
To further reduce servo shutdowns, we punished high-frequency gaits during evolution. We calculated a gait frequency as the average number of servo direction changes per leg per second. If this frequency was higher than an experimentally-selected threshold of 1.67 Hz, the measure of distance traveled by the center of mass during the gait was reduced exponentially by multiplying it by a discount factor, which was $e^{(1.67-\mathrm{freq})}$. Following Clune et al. 2009~\cite{clune2009evolving}, fitness equation was $2^{\mathrm{distance}^{2}}$. After evolving the gaits in simulation, the champion gait of the last generation of each of the 20 runs was transferred onto the real robot and the distance traveled was measured.

### 2012-10-30 paper_sean/simulator.tex

% !TEX root =main.tex
The simulator was the same as Glette et al. 2012 \cite{glette}. C++ code for the the Nvidia PhysX physics engine approximately describes the QuadraTot, including the mass and size of the robot components and its degrees of freedom~(\figref{quadratotWhiteBooties}). In the simulator, each individual joint range was limit as described above, but Smart Cropping was not included because it hindered performance by limiting the types of gaits evolution could explore in early generations.

>>>>

% !TEX root =main.tex

\subsubsection{Simulator:}
Gaits evolved in the simulator from Glette et al., which represents QuadraTot in the Nvidia PhysX physics engine\footnotemark[\value{footnote}], including the mass and size of the QuadraTot components and its degrees of freedom~(\figref{quadratotWhiteBooties}, right). In the simulator, each individual joint range was limit as described above, but Smart Cropping was not included because we found that it hindered performance by limiting the types of gaits evolution could explore in early generations.
